\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}

### 2014-06-09 paper/subdocuments/data_and_methods.Rnw
FTA data description + comparison table (FTA vs survey)

Studies on inequality have to address several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All this questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss  the advantages and shortcomings of tax data compared to other data sources - namely survey data. Along with the description we explain the methods and techniques we use to construct time-series of inequality-measure for income for Switzerland, which we will present in the result sections. By doing this, we demonstrate the suitability of the FTA Tax Data to assess the development of inequality in Switzerland

@@

@@

The issue of incomplete coverage is far less dramatical with tax data. Essentially every permanent resident in Switzerland, who has completed the age of 18 years (respectively 20 years prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). In principle this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover several unusual situations. Most important high net wealth individuals taxed according to their expenditures.However, special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income do not show up in the statistics if their income after deductions falls below 15'000 CHF and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units are reported, but not for the years before.

@@

@@

>>>>

Studies on inequality have to address several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All these questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss the advantages and shortcomings of tax data compared to other data sources - namely survey data. We describe and explain methods and techniques needed to construct time-series of income inequality measures for Switzerland. We will present these time-series in the result section and evaluate the use of the FTA Tax Data to assess the development of inequality in Switzerland

@@

Our data comes from the Swiss Federal Tax Administration (FTA). Federal taxes are collected and documented by the FTA since 1915. Being called a war-tax in the beginning, the federal tax was renamed to crisis levy in 1934, defense-tax in 1939 and is finally known as direct federal tax since 1983. The time frame we were able to collect ranges from 1945 to 2010 including 44 tax periodes. \footnote{Between 1993 and 2003 there is no exact data available for Switzerland because of a system change from taxation assessed in arrears (Praenumerando-System) to taxation assessed on current year income (Postnumerando-System) which was implemented by cantons in different years.}  While the FTA provides data in electronic form since 1973 we collected earlier data by scanning hard copies. Data is available for Switzerland plus all cantons and basically covers every tax unit (individual or household) in Switzerland liable to pay federal taxes. This exempts all tax units with taxable income below a certain threshold (e.g. 15.000 CHF in 2010). Furthermore the FTA differentiates between two groups of tax units, so called normal cases and special cases. A normal case is a tax unit residing in a swiss canton without foreign source income and being liable to taxation all year long. All other tax units and very few that are taxed based on the style of living because they don't work (Pauschalbesteuerte) are special cases. 
The FTA provides two income measures: taxable income and net income. Net income here is an administrative term and means taxable income plus social deductions (children and supported persons) but not including other deductions like donations or health-care costs. Both measures are designed for taxation purposes which might limit the suitability to measure inequality as we will discuss later.

@@

The issue of incomplete coverage is far less dramatical with tax data. Essentially every permanent resident in Switzerland, who has completed the age of 18 years (respectively 20 years prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). In principle this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover several unusual situations. Most important high net wealth individuals taxed according to their expenditures.However, special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income does not show up in the statistics if their income after deductions falls below 15'000 CHF and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units are reported, but not for the years before.

@@

\emph{Intertemporal comparison}

One advantage of tax data over survey data is their observation period. While the FTA data we use range back to 1945, the earliest period of survey data is 2000 (HABE data). To be fair, survey data might cover the late periods (2011 and 2012) which are not covered by tax data. The striking feature of the long FTA time series is its consistency. Both population coverage and income measures are consistent for the complete observation period. This can not be said for survey data as these are based on samples and therefore require an ideal sampling design or reweighting to be representative and comparable over time. We will see how well sampling and weighting is done in the results section.

@@

\begin{table}[h]
\begin{tabular}{lcccc}
                                   & FTA       & HABE      & EU-SILC   & LIS       \\
Range                              & 1945-2010 & 2000-2011 & 2007-2012 & 1982-2004 \\
Cross country comparison possible? & no        & no        & yes       & yes       \\
Top incomes covered?               & yes       & no        & no        & no        \\
Low incomes covered?               & no        & yes       & yes       & yes       \\
Concepts of economic ressources?   & no        & yes       & ?         & ?         \\
Equivalisation?                    & no        & yes       & ?         &          
\end{tabular}
\end{table}
